Supplementary Information

### 2021-08-10 doc/manuscript/response_to_reviews.Rmd

**The key point with the hypotheses referenced is that we hypothesize that "[RW] *initially* increases with DBH for shade-tolerant species." This hypothesis does not preclude a subsequent decline in RW across larger size classes, which indeed was the most common outcome (described in results, "Variation with DBH" section, paragraph starting with "For RW, DBH was included..."). We have carefully reviewed the language around this issue to ensure that descriptions of an "increasing" trend in RW refer only to the smaller end of the DBH range, and have found all of the language to be accurate. While a more thorough investigation of DBH trends in growth rate and their interrelationships across growth metrics could be interesting and informative, this would be beyond the scope of the current paper.**

**More broadly, we believe that the relationships among results for the various growth metrics are fundamentally consistent. We would expect, and observe, some modest differences in climate variables selected (e.g., because the various growth metrics would effectively place different weight on above- or below-average growth rates for large vs small trees), variables included in the best model, etc. However, the results across growth metrics are broadly consistent.**

 L222-224:
Dealing with tree and forest mortality is difficult. However, climate change is predicted to affect mortality in many forests. In turn, mortality can impact the growth of the remaining trees depending on the stature and location of the dead tree. This point should be addressed- the article presents the relationship between growth and climate only for the surviving trees. How does excluding growth data from the analysis potentially bias the results?

**This is a good point. Increased mortality would result in decreased competition and potential growth release for surviving individuals. Thus, in theory, we might expect increasing mortality rates in many forests around the world [@mcdowell_pervasive_2020] to result in trends of increasing growth. However, mortality trends may not be notable enough at our sites to produce this effect, or may even go in the opposite direction (e.g., reduced intermediate disturbances at CB). Moreover, increased mortality is often correlated with decreased growth unless the conditions reverse (so, for example, the trend at SC is towards increased mortality and decreased growth). Overall, in this analysis, negative growth trends predominate, and so we do not believe that growth release from increasing mortality is a dominant factor. However, we have added mention of changing mortality rates in the discussion: "Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table 1, Fig. 7), probably reflecting some combination of successional changes, *changing mortality rates*, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009]."**

@@

Line 545-547: Agree that this is useful to help with prediction, but what is missing from the article is a more clear idea of how this could/would be done. What emerges here is that trees and forests are complicated. There seem to be some general patterns emerge, but for each of these there are always exceptions. How do you propose synthesizing all of this to actually improve predictions?

**We agree that forests are complex, and this method both reveals and allows us to start to distill some of that complexity. We have added a summary figure (new Fig. 3) to help to highlight some of the general patterns. There is obviously a lot more work to be done on this theme that is beyond the scope of the current analysis.**

Section 590: It is really interesting to contract the variability between population scale and individual scale. There was a great paper by Jim Clark that constrasted these scales and showed that responses could appear different for each. Can’t remember the reference, but this seems to be the theme of this paper: individual scale can be quite different from population scale.

**Good point. To remind readers that our models refer to population-mean responses, we have inserted phrases such as "mean" or "on average" in appropriate places throughout the results and discussion. We have also added specific mention of the principle highlighted by the Clark paper: "In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011]."**

### 2021-08-10 doc/manuscript/ForestGEO_treerings.Rmd
finished responses to reviews!

Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied with DBH for most species at all sites (Fig. `r size`).
Because the effects of calendar year could not be evaluated for all species (Fig. `r SummaryFigure`), the DBH trends described here refer to models without year. 
Relationships between growth rate and DBH were best described by models with second-order terms for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`).

@@

The long-term growth records contained in tree rings provide an exceptional tool for understanding past drivers of growth and anticipating future forest changes, yet traditional dendrochronological analysis methods were not designed to disentangle multiple, simultaneously acting drivers of tree growth, nor their implications for whole-forest productivity.
Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions.
Analyzed with respect to each of  these drivers individually, our method yields results that are consistent with what would be obtained using established methods. 
Beyond this, because our approach considers these factors simultaneously, it allows analyses of their joint and interactive effects.
Applied across a wide range of forest types and species distributed globally across 10 sites, we have shown that tree species vary in the shapes of their functional responses with respect to size-related sensitivity to different climate variables.
Dissecting these species-specific long-term responses is essential to understanding the drivers of variability and directional changes in tree growth over the past century, and to predicting changes in forest composition and function in the future.

@@

There was substantial variation across species-site combinations in the relationship between DBH and growth rate (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, RW declined with DBH for a substantive portion of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
These results imply that while trees growing in high-light conditions typically display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].

Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

In all seven sites dominated by trees less than 100 years old, growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
While attribution of these trends is difficult, it is likely that some trends are caused by stand dynamics as cohorts and stands develop over time.
Such negative trends are fairly typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990]. 
For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004; @west_increasing_2020].
Gradual shifts in abiotic drivers (e.g., indirect effects of warming) likely also play a role at some of these sites.
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]. At this site, we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.

Within the three older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Fig. `r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s, consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008]. 
Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose growth rate increased significantly over the analysis time frame (Fig. `r year_fig`).

>>>>

Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied with DBH for most species at all sites (Fig. `r size`).
Because the effects of calendar year could not be evaluated for all species (Fig. `r SummaryFigure`), the DBH trends described here refer to models without year. 
Relationships between population-mean growth rate and DBH were best described by models with second-order terms for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`).

@@

The long-term growth records contained in tree rings provide an exceptional tool for understanding past drivers of growth and anticipating future forest changes, yet traditional dendrochronological analysis methods were not designed to disentangle multiple, simultaneously acting drivers of tree growth, nor their implications for whole-forest productivity.
Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions.
Analyzed with respect to each ofthese drivers individually, our method yields results that are consistent with what would be obtained using established methods. 
Beyond this, because our approach considers these factors simultaneously, it allows analyses of their joint and interactive effects.
Applied across a wide range of forest types and species distributed globally across 10 sites, we have shown that tree species vary in the shapes of their functional responses with respect to size-related sensitivity to different climate variables.
Dissecting these species-specific long-term responses is essential to understanding the drivers of variability and directional changes in tree growth over the past century, and to predicting changes in forest composition and function in the future.

@@

There was substantial variation across species-site combinations in the population mean relationship between DBH and growth rate (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
These results imply that while species that typically grow in high-light conditions commonly display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), average RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018]. 
Our finding that on average across populations, BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that mean $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

In all seven sites dominated by trees less than 100 years old, population-mean growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
While attribution of these trends is difficult, it is likely that some trends are caused by stand dynamics as cohorts and stands develop over time.
Such negative trends are fairly typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990]. 
For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004; @west_increasing_2020].
Gradual shifts in abiotic drivers (e.g., indirect effects of warming) likely also play a role at some of these sites.
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]. At this site, we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.

Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table `r hypotheses`, Fig. `r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s, consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008]. 
Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose population-mean growth rate increased significantly over the analysis time frame (Fig. `r year_fig`).
